---
title: "Visualizations in R using the tidyverse"
subtitle: "Sigma Xi Fall Workshop, 2023"
author: "P. Lombardo & J. Kaufman"
output:
  html_document:
    df_print: paged
  html_notebook: default
fig.width: 8
fig.height: 4
---

Begin by loading the necessary packages and necessary data sets.
```{r set-up, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(RColorBrewer)

# Gapminder data set, courtesy of the gapminder package
gapminder<-read.csv('data/gapminder.csv',header=T)

# Courtesy of Kaggle
spotify<-read.csv('data/spotify-2023.csv',header=T)
    # Select only certain variables
spotify<-spotify %>%
    mutate(danceability=danceability_.) %>%
    select(track_name,artist.s._name,
           released_year,
           in_apple_charts,
           danceability, 
           mode, streams)

# Couresty of ggplot maps and 
# World Happiness Report
world_happiness<-read.csv('data/worldHappiness.csv',
                          header=T)

# Courtesy of five-thirty-eight
drugUse<-read.csv('data/drug-use-by-age.csv',
                  header=T,
                  na.strings = "-")
```


# Workshop Examples

## Case Study 1: Incrementally building plots with the `iris` data set.
The `iris` data set is pre-loaded in `R`, and a very common data set for statistics, data science, and data visualization.

```{r}
head(iris)
```

Let's build our plot!
```{r}

```

Let's improve our plot!
```{r}

```

```{r}
# display.brewer.all()
```

Demonstrating Simpson's Paradox.
```{r}
# place code here
```


## Case Study 2: Using `spotify` data to explore plot options and `patchwork` them together

Here is our data set: [Spotify songs 2023 (Kaggle)](https://www.kaggle.com/datasets/nelgiriyewithana/top-spotify-songs-2023).

```{r}
head(spotify)
```


**Plot 1: Histogram Example**.
```{r}
# Let's build a histogram together
# p1
ggplot(data = spotify,
       aes(x = ... )+
    geom_histogram()
```

**Plot 2: Fancy Violin Plot Example**.
Let's look at popularity by adding a new variable, `aboveMedian`. This variable will have "Yes" if the song has more streams than the median number of streams, and "No" otherwise. 

Next, we'll plot a violin plot, the love-child of a box-plot and a histogram!

```{r}
# What's a violin plot?
# p2
spotify %>%
    mutate()
    # ggplot()+
    #     geom_violin()
    # Some more code


    # scale_color_brewer("Key Signature Mode",palette = "Set1")+
    #     labs(color="Key Signature Mode",
    #          y="Danceability Score",
    #          x="Total Streams Above Median")+
    #     theme_bw()+
    #     theme(legend.position="top")
```

**Plot 3: Bar Plot Example**.
```{r}
# how about a simple bar plot...
# p3
# ggplot(data = spotify,
#            aes())+
#     geom_bar()
# More code coming

    # scale_fill_brewer("Key Signature Mode",palette = "Set1",
    #                       guide = "none")+
    #     labs(x="Key Signature Mode",
    #          y="Percentage")+
    #     theme_bw()
```

Simple patchwork options:
```{r}
# (p1+p2)/p3
```

More advanced layouts:
```{r}
layout <- "
AAAA##
BBBBCC
BBBBCC
"
# p1+p2+p3 +plot_layout(design=layout) +
#     plot_annotation(title = "Spotify Plots")
```


## Case Study 3: Plotting summary data with the `gapminder` data set

The gapminder data set looks at metrics like life expectancy and national GDP for several nations around the world over time. The data here is borrowed directly from the [`gapminder` R package](https://cran.r-project.org/web/packages/gapminder/index.html), but more data sets are available at the [Gapminder organization](https://www.gapminder.org/) website.

Let's make a plot consider trends in life expectancy over time by continent.
```{r, warning=FALSE, message=FALSE}
head(gapminder)
```

We first need to make a data frame of summaries that give us our plotting elements:

* mean life expectancy (and standard deviation, with `low` whisker and `high` wisker if you want errorbars)
* mean population, measured in millions.

```{r}
gapminder %>%
    group_by(...) %>%
    summarize(...)
```

Let's make our plot!
```{r}
ggplot(data = gapminder_summary,
       aes(x = ...,
           y = ...,
           color=...,
           group=...))+
    geom_point()+
    geom_line()
    # More code below (geom_errobar and geom_ribbon for variation)


    # labs(x="Year",
    #      y="Mean Life\nExpectancy",
    #      title="Time trends in mean life expectancy by continent")+
    # theme_classic()

    # Futher theme elements can change this to "dark mode", e.g.
    # theme(text = element_text(color="white"),
    #       plot.title = element_text(size = 16),
    #       axis.text= element_text(color="white"),
    #       plot.background = element_rect(fill = "gray12"),
    #       legend.background = element_rect(fill ="gray12"),
    #       panel.background = element_rect(fill = "gray12",
    #                             colour = "gray",
    #                             size = 0.5, linetype = "solid"))
```


## Case Study 4: Map data and World Happiness
Part of our data comes from here: [World Happiness Report (data)](https://worldhappiness.report/data/).

We mixed this happiness report with latitude and longitude data from ggplot, specifically `ggplot2::map_data("world")`.

```{r}
head(world_happiness)
```

Let's start by building a world map first, then we'll color the countries according to there `Perceptions.of.corruption`.
```{r}
ggplot(data = world_happiness %>% arrange(order),
       aes(...))+
    geom_polygon(color="black") # adjust alpha and linewidth
    # More code here...

    # For continuous color or fill scales, use...
    # scale_fill_gradientn("Perception of\nCorruption",
    #                      colors = brewer.pal(...),
    #                      na.value=...)
    # theme_bw()+
    # labs(x="Longitude", y = "Latitude",
    #          title = "Perceptions of Corruption by Country",
    #          subtitle = "(Data provided by World happiness Report)")
```



## Case Study 5: Reshaping and plotting with the `drugUse` data set.

Here is our data set: [druge-use data set](https://github.com/fivethirtyeight/data/tree/master/drug-use-by-age)

```{r}
head(drugUse)
```

Columns display the same measurements (`use` as percentage of the age group that used in the 12-month period, and `frequency` as the median number of times an individual used in the 12-month period), but categories are split among columns.

One tidy format would have `use` and `frequency` as their own columns, with `drug_type` as indicating the type of drug.

Our steps will be:

1. Use pivot longer on all columns *except* `age` and `n`.
2. Use the `names_pattern` argument to split these column titles by drug type and measurement type ("use" versus "frequency")
3. Then we will pivot wider to put `use` and `frequency` as their own variables.

For those interested in the documentation:
```{r}
?pivot_longer
?pivot_wider
```

Let's build our data frame!
```{r}

```


```{r}
# Let's create a heatmap with geom_tile()
ggplot(data = ...,
       aes(...)) +
    geom_tile()
    # More code below

    # labs(x="Age Group",
         y="Drug Type",
         title="Drug-use Percentages by Age")
    # Can adjust outer background color to make the colors
    # easier to see
    # theme(...)

```

# Helpful Resources:

## Data Repositories

* [Kaggle data sets](https://www.kaggle.com/datasets)
    * [Spotify songs 2023 (Kaggle)](https://www.kaggle.com/datasets/nelgiriyewithana/top-spotify-songs-2023)
* [World Happiness Report (data)](https://worldhappiness.report/data/)
* [five-thirty-eight data (github)](https://github.com/fivethirtyeight/data)
    * [druge-use data set](https://github.com/fivethirtyeight/data/tree/master/drug-use-by-age)


## Package documentation

* [`tidyverse` documentation](https://www.tidyverse.org/)
    * [`ggplot` documentation](https://ggplot2.tidyverse.org/)
    * [`tidyr documentation`](https://tidyr.tidyverse.org/index.html)
    * [`dplyr` documentation](https://dplyr.tidyverse.org/)
* [`patchwork` documentation](https://patchwork.data-imaginist.com/)
* [`gapminder` R package documentation](https://cran.r-project.org/web/packages/gapminder/index.html)
    * [Gapminder organization](https://www.gapminder.org/)



