---
title: "Visualizations in R using the tidyverse"
subtitle: "Sigma Xi Fall Workshop, 2023 (With Code)"
author: "P. Lombardo & J. Kaufman"
output:
  html_document:
    df_print: paged
  html_notebook: default
fig.width: 8
fig.height: 4
---

Begin by loading the necessary packages and necessary data sets.
```{r set-up, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(RColorBrewer)
knitr::opts_chunk(fig.width=8, fig.height=6)

# Gapminder data set, courtesy of the gapminder package
gapminder<-read.csv('data/gapminder.csv',header=T)

# Courtesy of Kaggle
spotify<-read.csv('data/spotify-2023.csv',header=T)
    # Select only certain variables
spotify<-spotify %>%
    select(track_name,artist.s._name,
           released_year,
           in_apple_charts,
           danceability_., 
           mode, streams)

# Couresty of ggplot maps and 
# World Happiness Report
world_happiness<-read.csv('data/worldHappiness.csv',
                          header=T)

# Courtesy of five-thirty-eight
drugUse<-read.csv('data/drug-use-by-age.csv',
                  header=T,
                  na.strings = "-")
```


# Workshop Examples

## Case Study 1: Incrementally building plots with the `iris` data set.
```{r}
head(iris)
```

```{r}
ggplot(data = iris,
       aes(x = Petal.Width,
           y = Sepal.Width))+
    geom_point()+
    geom_smooth(method="lm", formula=y~x, se = F)+
    scale_y_continuous(limits=c(0,5))+
    labs(x="Petal Width (in)",
         y="Sepal Width (in)",
         title="Comparing Sepal and Petal Width")+
    theme_bw()
```

```{r}
ggplot(data = iris,
       aes(x = Petal.Width,
           y = Sepal.Width,
           color=Species))+
    geom_point()+
    geom_smooth(method="lm", formula=y~x, se = F)+
    scale_y_continuous(limits=c(0,5))+
    scale_color_brewer(palette="Set2")+
    labs(x="Petal Width (in)",
         y="Sepal Width (in)",
         title="Comparing Sepal and Petal Width By Species")+
    theme_bw() 
```

Piecing them together with `patchwork`
```{r}
ggplot(data = iris,
       aes(x = Petal.Width,
           y = Sepal.Width))+
    geom_point()+
    geom_smooth(method="lm", formula=y~x, se = F)+
    scale_y_continuous(limits=c(0,5))+
    labs(x="Petal Width (in)",
         y="Sepal Width (in)",
         title="Comparing Sepal and Petal Width")+
    theme_bw()->p1

ggplot(data = iris,
       aes(x = Petal.Width,
           y = Sepal.Width,
           color=Species))+
    geom_point()+
    geom_smooth(method="lm", formula=y~x, se = F)+
    scale_y_continuous(limits=c(0,5))+
    scale_color_brewer(palette="Set2")+
    labs(x="Petal Width (in)",
         y="Sepal Width (in)",
         title="Comparing Sepal and Petal Width By Species")+
    theme_bw() -> p2

p1+p2+plot_annotation(title="Simpson's Paradox",
                      theme= theme(plot.title = element_text(size = 18)))
```

## Case Study 2: Using `spotify` data to explore plot options and `patchwork` them together

```{r}
head(spotify)
```


**Plot 1: Histogram Example**.
```{r}
# Let's build a histogram together
# p1
ggplot(data = spotify,
       aes(x = danceability_.,
           fill = mode))+
    geom_histogram(breaks=seq(20,100,by=10),
                   position=position_dodge(width=2),
                   alpha=.5)+
    scale_x_continuous(breaks=seq(20,100,by=10))+
    scale_fill_brewer(palette = "Set1",
                      guide = "none")+
    labs(x="Danceability Score")+
    theme_bw()->p1
    
```

**Plot 2: Fancy Violin Plot Example**.
```{r}
# What's a violin plot?
# p2
spotify %>%
    arrange(desc(streams)) %>%
    mutate(aboveMedian=ifelse(streams > median(streams),"Yes","No")) %>%
    ggplot(data = .,
           aes(x = aboveMedian,
               y = danceability_.))+
    geom_violin(fill='gray')+
    geom_point(position=position_jitter(width=.15),
               alpha=.4,
               aes(color=mode))+
    scale_color_brewer("Key Signature Mode",palette = "Set1")+
    labs(color="Key Signature Mode",
         y="Danceability Score",
         x="Total Streams Above Median")+
    theme_bw()+
    theme(legend.position="top")-> p2
```

**Plot 3: Bar Plot Example**.
```{r}
# how about a simple bar plot...
# p3
spotify %>%
    arrange(desc(streams)) %>%
    mutate(aboveMedian=ifelse(streams > median(streams),"Yes","No")) %>%
    ggplot(data = .,
           aes(x = mode,
               y = after_stat(count)/sum(after_stat(count)),
               fill=mode))+
    geom_bar(alpha= .7)+
    scale_y_continuous(labels=scales::percent)+
    scale_fill_brewer("Key Signature Mode",palette = "Set1",
                      guide = "none")+
    labs(x="Key Signature Mode",
         y="Percentage")+
    theme_bw()->p3
```

Simple patchwork options:
```{r}
(p1+p2)/p3
```

More advanced layouts:
```{r}
layout <- "
AAAA##
BBBBCC
BBBBCC
"
p1+p2+p3 +plot_layout(design=layout)+
    plot_annotation(title = "Spotify Plots")
```


## Case Study 3: Plotting summary data with the `gapminder` data set

The gapminder data set looks at metrics like life expectancy and national GDP for several nations around the world over time. 

Let's make a plot consider trends in life expectancy over time by continent.
```{r, warning=FALSE, message=FALSE}
head(gapminder)
```

Summarize first.
```{r}
gapminder %>%
    group_by(continent,year) %>%
    summarize(meanLife = mean(lifeExp),
              sdLife = sd(lifeExp),
              low = meanLife - 2*sdLife,
              high = meanLife +2*sdLife,
              meanPop = mean(pop)/1000000) -> gapminder_summary
```

Make your plot.
```{r}
ggplot(data = gapminder_summary,
       aes(x = year,
           y = meanLife,
           color=continent,
           group=continent))+
    geom_point(aes(size = meanPop))+
    geom_line()+
    scale_color_brewer("Continent",palette="Set2")+
    # geom_errorbar(aes(ymin=low,ymax=high),
    #               alpha=.3,width=1)+
    # geom_errorbar(aes(ymin=low,ymax=high),
    #               alpha=.3,width=1,
    #                 position=position_dodge())+
    # or use ribbons to show variation
    # geom_ribbon(aes(ymin=low,ymax=high,fill=continent),
    # alpha=.1)+
    scale_fill_brewer("Continent",palette="Set2", guide = "none")+
    scale_x_continuous(breaks=seq(1950,2020,by=10))+
    scale_size_continuous("Mean Population\nin Millions")+
    geom_hline(yintercept=seq(40,80,by=10),
               col='gray',alpha=.3)+
    labs(x="Year",
         y="Mean Life\nExpectancy",
         title="Time trends in mean life expectancy by continent")+
    theme_classic()
```

### Case Study 3b: Map data and World Happiness
```{r}
world_map<-ggplot2::map_data("world")
```

```{r}
# Let's create a map with world_happiness
ggplot()+
    geom_polygon(data =world_map,
       aes(x=long,y=lat,group=group),color='black',
       fill='gray',alpha=.6)+
# Let's color the map by `Perceptions of Corruption`
    coord_cartesian(ylim=c(-55,90))+
    theme_bw()+
    geom_polygon(data = world_happiness %>% arrange(order),
                 aes(x=long,y=lat,group=group,
                     fill=Perceptions.of.corruption))+
    scale_fill_gradientn("Perception of\nCorruption",colors = rev(brewer.pal(9,name="RdYlBu")),
                         na.value='black')+
    labs(x="Longitude", y = "Latitude",
         title = "Perceptions of Corruption by Country",
         subtitle = "(Data provided by World happiness Report)",
         caption = "Note: Antarctica has been removed; penguins are quite happy.")
```



## Case Study 4: Reshaping and plotting with the `drugUse` data set.

```{r}
head(drugUse)
```

Columns display the same measurements (`use` as percentage of the age group that used in the 12-month period, and `frequency` as the median number of times an individual used in the 12-month period), but categories are split among columns.

One tidy format would have `use` and `frequency` as their own columns, with `drug_type` as indicating the type of drug.

```{r}
?pivot_longer
?pivot_wider
```

```{r}
# display.brewer.all()
```

```{r}
# Let's create a heatmap with geom_tile()
```

# Helpful Resources:

## Data Repositories

* [Kaggle data sets](https://www.kaggle.com/datasets)
    * [Spotify songs 2023 (Kaggle)](https://www.kaggle.com/datasets/nelgiriyewithana/top-spotify-songs-2023)
* [five-thirty-eight data (github)](https://github.com/fivethirtyeight/data)
    * [druge-use data set](https://github.com/fivethirtyeight/data/tree/master/drug-use-by-age)

## Package documentation

* [`tidyverse` documentation](https://www.tidyverse.org/)
    * [`ggplot` documentation](https://ggplot2.tidyverse.org/)
    * [`tidyr documentation`](https://tidyr.tidyverse.org/index.html)
    * [`dplyr` documentation](https://dplyr.tidyverse.org/)
* [`patchwork` documentation](https://patchwork.data-imaginist.com/)
* [`gapminder` R package documentation](https://cran.r-project.org/web/packages/gapminder/index.html)
    * [Gapminder organization](https://www.gapminder.org/)
* [World Happiness Report (data)](https://worldhappiness.report/data/)


